Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework

### 2016-07-07 ms/ms.Rnw
Edited intro. rewrote results. Need to add a couple of numbers and references in places

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Tropical trees $|$ wood density $|$ growth $|$ mortality $|$ interspecific}

% 250 words
\begin{abstract}
Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects. In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
@

@@

\dropcap{P}lant mortality rates are known to vary among individuals within a species, among coexisting species within the same forest, from year-to-year, and between forests [refs]. Current estimates suggest the range of variation is large, e.g. from xxx xxx among plants within the same forest, and from XX-XXX among forests within any given year, or between years. This variation has considerable consequences for forest structure and dynamics. Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}. Improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.

Two substantial challenges exist when studying the mortality process in trees. The first is that very large population sizes and/or very long periods of observation are needed for effective inference. This requirement arises is in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are needed, e.g. 100s-1000s records. As such, detailed studying of mortality for forest trees has only recently become possible, with the accumulation of data from repeat surveys in long-term monitoring plots containing 1000's trees and spanning several decades \cite{Anderson-Teixeira-2015a}.

\begin{figure}[ht]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures on stem diameter ($D$) and status ($S$, alive or dead) for individual tress at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects. The parameters of the model are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes ingrowth rate; and $\gamma$ is the asymptote, or baseline hazard. Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g. windfall, fire) that kill a plant irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Log-loss function
D) The predictive skill of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
\label{fig1}
\end{figure}

The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX). Generally, however, these influences have not been formalised as hazard functions. Yet such knowledge is needed for predicting mortality into the future and in models of ecosystem function (XX). For projecting the outcomes of a population, hazard functions are invaluable because they ....

A particular challenge is to integrate the apparently conflicting relationships of mortality to growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species also having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality into a unified model, including individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework for 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. To derive this unified model, we compare the skill of three alternative functional forms for $\lambda_i(t)$ in predicting patterns of tropical tree mortality in a diverse tropical forest in Panama (Fig. \ref{fig1}B). The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (refs). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific hypotheses:
\begin{enumerate}\itemsep 0pt
  \item Does mortality vary substantially between censuses?
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of total inter-specific variability explained by wood density?
\end{enumerate}
Answering these questions involves comparing predictive skill of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the five hypotheses evaluated above, and gradually increasing model complexity. At each stage, the best model from the previous stage was taken as input for the next stage. for example, stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig. \ref{fig1}B) using either annual diameter increment or area growth rates. In this stage each model was fitted without allowing parameters to vary by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. Following this, we used the best model of stage 2 but allowed its parameters to vary by species. This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.

@@

\subsection*{Predictive capacity of functional form and growth measure}
Using 10-fold cross validation our first aim was to determine which model form (Fig. \ref{fig1}b) and growth measure (i.e. DBH growth or basal area growth) best predicted tree mortality. We found that the model which accounted for both baseline (i.e. growth-independent) and growth-dependent hazards had a significantly lower logarithmic loss, and thus, a higher predictive capacity, relative to either the baseline only or growth-dependent only model variants (Fig. \ref{fig2}a). However, this improvement was dependent on growth measure, with predictive capacity increasing when DBH growth was used but not when basal area growth was used (Fig. \ref{fig2}a).

\begin{figure}[ht]
\includegraphics{figures/fig2.pdf}
\caption{Average logarithmic loss (lower = higher accuracy). A) Four sequential stages of model selection. Null model: shows the expected logloss for the null model with no census effect; because mortality varied substantially between censuses it was included on all subsequent models; Functional form: comparison between baseline hazard model ($\gamma c$; circle), growth\textendash dependent hazard model ($(\alpha e^{\beta X_{i}}) c$; triangle) and model incorporating both baseline and growth\textendash dependent hazards ($(\alpha e^{\beta X_{i}} + \gamma) c$; square); Wood density: Most parsimonious effect of wood density; Species effects: Model with species random effects on all parameters \textendash indicating the total gain in predictive capacity if accounting for all species variability. B) Wood density effects on all parameter combinations in the model that incorporates both baseline and growth-dependent hazards. Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. Errors are $\pm$ 95\% BCI.}
\label{fig2}
\end{figure}

\subsection*{Predictive capacity from including wood density}
Using the model with both baseline and growth-dependent hazards coupled with DBH growth we then examined: 1) whether predictive capacity increased if parameters varied as a function of a species' wood density; 2) whether the effect of wood density was predominately on growth-dependent or baseline (growth-independent) hazard parameters; and 3) what proportional gain is achieved from including wood density relative to a full species random effect model that implicitly captures all species level variation. We found that the inclusion of wood density on any combination of the three parameters significantly improved predictive performance (i.e. lowered logarithmic loss) relative to a model without wood density effects (Fig. \ref{fig2}b). Furthermore, the greatest gain in predictive accuracy occurred when we allowed the baseline hazard $\gamma$ to vary by wood density, which accounted for \Sexpr{logloss_comparisons(logloss_summaries)[['rho_v_re']]}\% of the maximum predictive gain achievable in the fall species random effect model (Fig. \ref{fig2}a).

\begin{figure}[ht]
\includegraphics{figures/fig3.pdf}
\caption{Estimated relationship between wood density and baseline mortality rate. Points are species estimates and solid line is the mean species trendline. All errors are $\pm$ 95\% BCI.}
\label{fig3}
\end{figure}

\subsection*{Effects of wood density and time}
Once determining that the most predictive model was one that accounted for both growth-dependent and baseline hazards, used DBH growth, and allowed $\gamma$ to vary by wood density, we then fitted two final models to the entire dataset. The first contained both species random effects on all parameters and wood density on the baseline hazard, $\gamma$. This allowed us to examine the range of species growth-mortality curves, the effect of wood density and how mortality varied over time. We also fitted a second model that only contained species random effects. This allowed us to determine which parameter varied most by species and what proportion of the baseline hazard species variability was explained by wood density. We found that baseline hazard rates were significantly lower for species with high wood density relative to those with low wood density (Fig. \ref{fig3}). That is, high wood density species had a lower probability of dying irrespective of growth rate. For example, our model predicts that on average a species with a wood density of 0.3 has a instantaneous mortality rate of XX, whereas a species with a wood density of 0.8 has a instantaneous mortality rate of XX, translating to an annual mortality rate of XX\% and XX\%, respectively. Furthermore, our model also revealed that instantaneous mortality rates varied over time. Specifically, mortality rates at BCI decreased over time with hazard rates being scaled by \Sexpr{census_effects(final_model,1)},\Sexpr{census_effects(final_model,2)} and \Sexpr{census_effects(final_model,3)} for census periods, 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010, respectively.

\begin{figure}[ht]
\includegraphics{figures/fig4.pdf}
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig4}
\end{figure}

\subsection*{Sources of variation}
We found that the baseline hazard, $\gamma$, varied the most between species followed by $\alpha$ and least on $\beta$. We also found that the wood density effect on $\gamma$ explained approximately XX\% of the species variability on the baseline hazard.

\begin{figure}[ht]
\includegraphics{figures/fig5.pdf}
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig5}
\end{figure}

\begin{figure}[ht]
\includegraphics{figures/fig6.pdf}
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig6}
\end{figure}

@@

>>>>

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Bayesian $|$ Growth rates $|$ Hazard rates $|$ Tropical trees $|$ Wood density}

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and plot-level effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama. We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

@@

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs]. This variation has considerable consequences for forest structure and dynamics. Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools (Chave et al 200X), while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}. As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.

Two substantial challenges exist when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates. This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades \cite{Anderson-Teixeira-2015a}.

The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX). Generally, however, these influences have not been formalised as hazard functions (XX). Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function (XX). This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, alive or dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards. The parameters of the models are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote, or baseline hazard. Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g. windfall, fire) that kill a plant irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Log-loss function highlighting that the further the predicted probability is from truth the higher the logloss.
D) The predictive accuracy of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the estimated log loss for each of these 10 model fits.}
\label{fig1}
\end{figure}

In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.

Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does mortality vary substantially between censuses?
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\subsection*{Mortality strongly varies over time}
Tree mortality progressively decreased over time. Between 1995 and 2010, the highest proportion of trees dying between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive accuracy (i.e. lower logarithmic loss; Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

\begin{figure}[!h]
\includegraphics{figures/fig2.pdf}
\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive accuracy. 
A) Five sequential stages of model selection with each stage increasing in complexity. Null: null model ($\gamma$); Census: inclusion of census effects ($\delta$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{\beta X_{i}})\delta$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{\beta X_{i}} + \gamma)\delta$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Change in logloss under all wood density\textendash parameter combinations. Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square). Note: errors are too small to see in panel A.}
\label{fig2}
\end{figure}

\subsection*{Growth hazard functions}
We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e. $(\alpha e^{\beta X_{i}} + \gamma)\delta$) and used stem diameter growth rather than basal-area growth, significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e. $(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig. \ref{fig2}A). Surprisingly, we found that the growth-dependent only function did not improve predictive capacity over the baseline only function. Furthermore, we found that while the hazard function that incorporates both baseline and growth-dependent terms significantly improved model accuracy, this was only apparent if dbh growth was used.

\begin{figure}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Relationship between species baseline mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure}

\subsection*{Effect of wood density on mortality rates}
The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig. \ref{fig2} A,B). The most parsimonious combination that provided the largest gain in predictive accuracy (i.e. lowest logarithmic loss) occurred when only the baseline hazard term ($\gamma$) was varied as a function of species wood density ($\rho$; Fig. \ref{fig2} B). We found that wood density was negatively correlated with species baseline mortality rate (Fig. \ref{fig3}). This meant that species with low wood density had on average higher growth-independent mortality rates ((Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to high wood density species. While, wood density did improve model accuracy it only accounted for XX\% of the total gain in predictive capacity that could be obtained by accounting for species differences.

\subsection*{Proportion of variation explained}
Predicted mortality rates were most variable between species accounting for approximately XX\% followed by between census periods (accounting for XX\%). Wood density, a species level trait, accounted for approximately XX\% of the total species variability, suggesting that other species differences contribute to the ramaining XX\% unexplained species variation.

As expected, the growth-dependent hazard terms (i.e. the part of the model allowed to vary by individual growth) accounted for XX\% of the overall variation in mortality rates while the baseline (growth-independent) hazard accounted for XX\%.

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in instantaneous mortality rates (A,B), and its derivative, annual mortality probability (C,D), with increasing growth rates. A,C) Estimated species curves for all 203 BCI species used in this study. B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
\label{fig4}
\end{figure}

\subsection*{Mortality rates, species light requirements and species maximum dbh}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species that predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth stochastic chance and required faster growth rates for a given mortality rate relative to a species that readily recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

\begin{figure}[h!]
\includegraphics{figures/fig5.pdf}
\caption{Proportion of predicted variation explained.
A) Variation explained by wood density (grey) as well as species and census random effects.
B) Variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
\label{fig5}
\end{figure}

@@

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

\begin{figure}[h!]
\includegraphics{figures/fig6.pdf}
\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (A,C,E,G) and species maximum DBH (B,D,F,H). Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study. Blue line shows the average trendline as determined from standard linear regression}
\label{fig6}
\end{figure}
